---
title: An R Markdown document converted from "02_bulk_preprocessed.ipynb"
output: html_document
---

```{r}
options(jupyter.plot_mimetypes = 'image/png')
options(repr.plot.width = 1, repr.plot.height = 0.75, repr.plot.res = 100)
```

```{r}
library(readr)
library(DESeq2)
library(ggplot2)
library(cowplot)
```

# Downloading preprocessed samples

We first must download the preprocessed data from GEO!

Download the supplementary data from the following GEO samples and place the downloaded files into the bulk folder of the workshop material:

### wild-type cells samples

#### 24h samples
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3763467 #SUM159_DMSO_24h_R1

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3763468 #SUM159_DMSO_24h_R2

#### 3h samples
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3763469 #SUM159_DMSO_3h_R1

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3763470 #SUM159_DMSO_3h_R2

### resistant cells samples

#### 24h samples
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3763477 #SUM159_JQ1R_DMSO_24h_R1

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3763478 #SUM159_JQ1R_DMSO_24h_R2

#### 3h samples
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3763479 #SUM159_JQ1R_DMSO_3h_R1

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3763480 #SUM159_JQ1R_DMSO_3h_R2

# Analysis of 24h samples

Let's start with reading in the 24h samples

```{r}
SUM159_24h_r1 = read_tsv("../bulk/GSM3763467_SUM159_DMSO_24h_R1.counts.txt.gz", col_names=F)
SUM159_24h_r2 = read_tsv("../bulk/GSM3763468_SUM159_DMSO_24h_R2.counts.txt.gz", col_names=F)

SUM159jq1r_24h_r1 = read_tsv("../bulk/GSM3763477_SUM159_JQ1R_DMSO_24h_R1.counts.txt.gz", col_names=F)
SUM159jq1r_24h_r2 = read_tsv("../bulk/GSM3763478_SUM159_JQ1R_DMSO_24h_R2.counts.txt.gz", col_names=F)
```

And continue with the 3h samples:

```{r}
SUM159_3h_r1 = read_tsv("../bulk/GSM3763469_SUM159_DMSO_3h_R1.counts.txt.gz", col_names=F)
SUM159_3h_r2 = read_tsv("../bulk/GSM3763470_SUM159_DMSO_3h_R2.counts.txt.gz", col_names=F)

SUM159jq1r_3h_r1 = read_tsv("../bulk/GSM3763479_SUM159_JQ1R_DMSO_3h_R1.counts.txt.gz", col_names=F)
SUM159jq1r_3h_r2 = read_tsv("../bulk/GSM3763480_SUM159_JQ1R_DMSO_3h_R2.counts.txt.gz", col_names=F)
```

```{r}
colnames(SUM159_24h_r1) = c("gene", "SUM159_24h_r1")
colnames(SUM159_24h_r2) = c("gene", "SUM159_24h_r2")
colnames(SUM159jq1r_24h_r1) = c("gene", "SUM159jq1r_24h_r1")
colnames(SUM159jq1r_24h_r2) = c("gene", "SUM159jq1r_24h_r2")
```

```{r}
colnames(SUM159_3h_r1) = c("gene", "SUM159_3h_r1")
colnames(SUM159_3h_r2) = c("gene", "SUM159_3h_r2")
colnames(SUM159jq1r_3h_r1) = c("gene", "SUM159jq1r_3h_r1")
colnames(SUM159jq1r_3h_r2) = c("gene", "SUM159jq1r_3h_r2")
```

```{r}
sum159_24h.expr_df <- merge(SUM159_24h_r1, SUM159_24h_r2, by = 'gene')
sum159jq1r_24h.expr_df <- merge(SUM159jq1r_24h_r1, SUM159jq1r_24h_r2, by = 'gene')

sum159_3h.expr_df <- merge(SUM159_3h_r1, SUM159_3h_r2, by = 'gene')
sum159jq1r_3h.expr_df <- merge(SUM159jq1r_3h_r1, SUM159jq1r_3h_r2, by = 'gene')
```

```{r}
head(sum159_24h.expr_df)
```

```{r}
merge_24h.expr_df <- merge(sum159_24h.expr_df, sum159jq1r_24h.expr_df, by = 'gene')
merge_3h.expr_df <- merge(sum159_3h.expr_df, sum159jq1r_3h.expr_df, by = 'gene')
```

```{r}
head(merge_24h.expr_df)
```

```{r}
expr_df_3h = merge_3h.expr_df[grep("^__", merge_3h.expr_df$gene, invert=T),]
rownames(expr_df_3h) = expr_df_3h$gene
expr_df_3h$gene = NULL


expr_df_24h = merge_24h.expr_df[grep("^__", merge_24h.expr_df$gene, invert=T),]
rownames(expr_df_24h) = expr_df_24h$gene
expr_df_24h$gene = NULL
expr_df_24h
```

```{r}
pdat_24h = data.frame( id=c("SUM159_24h_r1","SUM159_24h_r2", "SUM159jq1r_24h_r1","SUM159jq1r_24h_r2"),
                       group=c("SUM159_24h", "SUM159_24h", "SUM159jq1r_24h", "SUM159jq1r_24h") )
pdat_24h
```

```{r}
pdat_3h = data.frame( id=c("SUM159_3h_r1","SUM159_3h_r2", "SUM159jq1r_3h_r1","SUM159jq1r_3h_r2"), group=c("SUM159_3h", "SUM159_3h", "SUM159jq1r_3h", "SUM159jq1r_3h") )
pdat_3h
```

## DESeq2 analysis for 24h

```{r}
dds_24h <- DESeqDataSetFromMatrix(countData=expr_df_24h, 
                                  colData=pdat_24h,
                                  design=~group, tidy = FALSE)
```

```{r}
dds_24h
```

```{r}
dds_24h <- DESeq(dds_24h)
res_lfc_24h <- lfcShrink(dds_24h, coef=resultsNames(dds_24h)[[2]], type="apeglm")
res_lfc_24h <- res_lfc_24h[order(res_lfc_24h$padj, decreasing=FALSE),]


res_24h <- results(dds_24h)
res_24h <- res_24h[order(res_24h$padj, decreasing=FALSE),]
```

```{r}
res_24h
```

```{r}
res_lfc_24h
```

```{r}
options(jupyter.plot_mimetypes = 'image/png')
options(repr.plot.width = 15, repr.plot.height = 10, repr.plot.res = 150)
```

```{r}
plotPCA( DESeqTransform( dds_24h ), intgroup="group" )
```

```{r}
head(dispersions(dds_24h))
```

```{r}
sizeFactors(dds_24h)
```

```{r}
dispersionFunction(dds_24h)
```

```{r}
plotDispEsts(dds_24h)
```

```{r}
allplots = list()

for (geneName in rownames(head(res_lfc_24h, n=6)))
{
    data = plotCounts(dds_24h, gene=geneName, intgroup="group",returnData = TRUE)

    p = ggplot(data, aes(x=group, y=count, color=group)) + 
  geom_jitter(size=3) + ggtitle(geneName)+
  scale_y_log10()

  allplots[[geneName]] = p
}

cowplot::plot_grid(plotlist= allplots)
```

```{r}
plotMA(res_24h, ylim=c(-2,2))
```

```{r}
plotMA(res_lfc_24h, ylim=c(-2,2))
```

```{r}
resultsNames(dds_24h)
```

```{r}
resLFC_24h <- lfcShrink(dds_24h, coef="group_SUM159jq1r_24h_vs_SUM159_24h", type="apeglm")
resNorm_24h <- lfcShrink(dds_24h, coef="group_SUM159jq1r_24h_vs_SUM159_24h", type="normal")
resAsh_24h <- lfcShrink(dds_24h, coef="group_SUM159jq1r_24h_vs_SUM159_24h", type="ashr")
```

```{r}
options(repr.plot.width = 15, repr.plot.height = 5, repr.plot.res = 150)

pltXLim = 10 ** ceiling(log10(max(counts(dds_24h, normalized=T))))
par(mfrow=c(1,4), mar=c(4,4,2,1))
xlim <- c(1,pltXLim); ylim <- c(-3,3)
plotMA(res_24h, xlim=xlim, ylim=ylim, main="No Shrink")
plotMA(resLFC_24h, xlim=xlim, ylim=ylim, main="apeglm")
plotMA(resNorm_24h, xlim=xlim, ylim=ylim, main="normal")
plotMA(resAsh_24h, xlim=xlim, ylim=ylim, main="ashr")

options(repr.plot.width = 15, repr.plot.height = 10, repr.plot.res = 150)
```

```{r}
outres_24h = res_lfc_24h[! is.na(res_lfc_24h$padj),]

#fetches normalized counts
normExpr_24h = counts(dds_24h, normalized=T)
cn_24h = colnames(normExpr_24h)
normExpr2_24h = cbind(rownames(normExpr_24h), normExpr_24h)
colnames(normExpr2_24h) = c("Geneid", cn_24h)

finalres_24h = data.frame(PROBEID=rownames(outres_24h), FC=outres_24h$log2FoldChange, PVAL=outres_24h$pvalue, ADJ.PVAL=outres_24h$padj)

write.table(finalres_24h, file="../bulk/24h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")
write.table(normExpr2_24h, file="../bulk/normed_expr.24h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")

printres_24h = finalres_24h[finalres_24h$ADJ.PVAL < 0.001, ]
write.table(printres_24h, file="../bulk/sig.24h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")

printres_24h.up = printres_24h[printres_24h$FC > 1, ]
write.table(printres_24h.up , file="../bulk/sig.up.24h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")

printres_24h.down = printres_24h[printres_24h$FC < -1, ]
write.table(printres_24h.down , file="../bulk/sig.down.24h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")

```

```{r}
top24h.de = finalres_24h[order(abs(finalres_24h$FC),decreasing=TRUE),]
top24h.de = top24h.de[top24h.de$ADJ.PVAL < 0.05, ]

top24h.de.500.down = head(top24h.de[top24h.de$FC < -1,], n=500)
top24h.de.500.up = head(top24h.de[top24h.de$FC > 1,], n=500)

write.table(top24h.de.500.down , file="../bulk/top.down.24h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")
write.table(top24h.de.500.up , file="../bulk/top.up.24h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")

write.table(top24h.de , file="../bulk/top.all.24h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")
```

```{r}
top24h.de
```

```{r}
library(ggrepel)
library(EnhancedVolcano)
```

```{r}
head(outres_24h)
```

```{r}
EnhancedVolcano(outres_24h,
    lab = rownames(outres_24h),
    x = 'log2FoldChange',
    y = 'pvalue')
```

```{r}
library(pheatmap)
library(RColorBrewer)
```

```{r}
pdat_24h
```

```{r}
sigExprData = normExpr_24h[head(rownames(outres_24h), n=50), ]

heat_colors <- brewer.pal(9, "YlOrRd")

annotation <- data.frame(pdat_24h$group)
rownames(annotation) = pdat_24h$id

grpColors <- c("darkgreen", "darkred")
names(grpColors) <- c("SUM159_24h", "SUM159jq1r_24h")
anno_colors <- list("pdat_24h.group" = grpColors)

pheatmap(sigExprData,
        color = heat_colors,
        cluster_rows = T,
        show_rownames = T,
        scale = "row", annotation=annotation, annotation_colors = anno_colors
        )
```

# Analysis of 3h samples

```{r}
dds_3h <- DESeqDataSetFromMatrix(countData=expr_df_3h, 
                                  colData=pdat_3h,
                                  design=~group, tidy = FALSE)

dds_3h
```

```{r}
dds_3h <- DESeq(dds_3h)
res_lfc_3h <- lfcShrink(dds_3h, coef=resultsNames(dds_3h)[[2]], type="apeglm")
res_lfc_3h <- res_lfc_3h[order(res_lfc_3h$padj, decreasing=FALSE),]


res_3h <- results(dds_3h)
res_3h <- res_3h[order(res_3h$padj, decreasing=FALSE),]
```

```{r}
res_3h
```

```{r}
res_lfc_3h
```

```{r}
plotDispEsts(dds_3h)
```

```{r}
allplots = list()

for (geneName in rownames(head(res_lfc_3h, n=6)))
{
    data = plotCounts(dds_3h, gene=geneName, intgroup="group",returnData = TRUE)

    p = ggplot(data, aes(x=group, y=count, color=group)) + 
    geom_jitter(size=3) + ggtitle(geneName)+
    scale_y_log10()

    allplots[[geneName]] = p
}

cowplot::plot_grid(plotlist= allplots)
```

```{r}
plotMA(res_3h, ylim=c(-2,2))
```

```{r}
plotMA(res_lfc_3h, ylim=c(-2,2))
```

```{r}
resLFC_3h <- lfcShrink(dds_3h, coef="group_SUM159jq1r_3h_vs_SUM159_3h", type="apeglm")
resNorm_3h <- lfcShrink(dds_3h, coef="group_SUM159jq1r_3h_vs_SUM159_3h", type="normal")
resAsh_3h <- lfcShrink(dds_3h, coef="group_SUM159jq1r_3h_vs_SUM159_3h", type="ashr")
```

```{r}
options(repr.plot.width = 15, repr.plot.height = 5, repr.plot.res = 150)

pltXLim = 10 ** ceiling(log10(max(counts(dds_24h, normalized=T))))
par(mfrow=c(1,4), mar=c(4,4,2,1))
xlim <- c(1,pltXLim); ylim <- c(-3,3)
plotMA(res_3h, xlim=xlim, ylim=ylim, main="No Shrink")
plotMA(resLFC_3h, xlim=xlim, ylim=ylim, main="apeglm")
plotMA(resNorm_3h, xlim=xlim, ylim=ylim, main="normal")
plotMA(resAsh_3h, xlim=xlim, ylim=ylim, main="ashr")

options(repr.plot.width = 15, repr.plot.height = 10, repr.plot.res = 150)
```

```{r}

outres_3h = res_lfc_3h[! is.na(res_lfc_3h$padj),]

normExpr_3h = counts(dds_3h, normalized=T)
cn_3h = colnames(normExpr_3h)
normExpr2_3h = cbind(rownames(normExpr_3h), normExpr_3h)
colnames(normExpr2_3h) = c("Geneid", cn_3h)

finalres_3h = data.frame(PROBEID=rownames(outres_3h), FC=outres_3h$log2FoldChange, PVAL=outres_3h$pvalue, ADJ.PVAL=outres_3h$padj)

write.table(finalres_3h, file="../bulk/3h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")
write.table(normExpr2_3h, file="../bulk/normed_expr.3h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")

printres_3h = finalres_3h[finalres_3h$ADJ.PVAL < 0.001, ]
write.table(printres_3h, file="../bulk/sig.3h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")

printres_3h.up = printres_3h[printres_3h$FC > 1, ]
write.table(printres_3h.up , file="../bulk/sig.up.3h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")

printres_3h.down = printres_3h[printres_3h$FC < -1, ]
write.table(printres_3h.down , file="../bulk/sig.down.3h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")
```

```{r}
top3h.de = printres_3h[order(abs(printres_3h$FC),decreasing=TRUE),]
top3h.de = top3h.de[top3h.de$padj < 0.05, ]


top3h.de.500.down = head(top3h.de[top3h.de$FC < -1,], n=500)
top3h.de.500.up = head(top3h.de[top3h.de$FC > 1,], n=500)

write.table(top3h.de.500.down , file="../bulk/top.down.3h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")
write.table(top3h.de.500.up , file="../bulk/top.up.3h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")


write.table(top3h.de , file="../bulk/top.all.3h_sum159_vs_sum159_jq1r.tsv", row.names=F, quote=F, sep="\t")
```

```{r}
EnhancedVolcano(outres_3h,
    lab = rownames(outres_3h),
    x = 'log2FoldChange',
    y = 'pvalue')
```

# Enrichment analysis on 24h samples

```{r}
library(disgenet2r)

Sys.setenv(DISGENET_API_KEY= get_disgenet_api_key(email="joppich@bio.ifi.lmu.de", password="disgenetpassword"))
```

```{r}
length(intersect(c("AREG","CCND1","BCL2","KLF5","CHKA","EPHB2","EREG","FYN","GLI2","GSN","FOXA1","CXCL8","KDR","MAPT","MMP1","RAB27B","SLC20A1","TRPM2","FOSL1","STC2","MAP4K4","ADAMTS2","SEMA4D","TRIM29","SGK3","PAG1","LGR6","UBASH3B","DNER","E2F7","CD24"),
head(top24h.de.500.up$PROBEID	, n=500)))
```

```{r}
input_genes = head(top24h.de.500.up$PROBEID, n=500)

print(input_genes)

disgenet_input <- gene2disease( gene = input_genes, vocabulary = "HGNC", database = "ALL")

common_genes = intersect(disgenet_input@qresult$gene_symbol, input_genes)

print(paste("Remaining Gene Count: ", length(common_genes), "of", length(input_genes)))

res_enrich <-disease_enrichment( entities =common_genes, vocabulary = "HGNC", database = "ALL")
```

```{r}
res_enrich@qresult[ res_enrich@qresult$FDR < 0.0001, c('ID','Description','source','Ratio','BgRatio','pvalue','FDR')]
```

```{r}
plot( res_enrich, class = "Enrichment",  cutoff= 0.05, nchars=70)
```

```{r}
library(clusterProfiler)
library(DOSE)
library(enrichplot)
library(org.Hs.eg.db)
```

```{r}
dim(finalres_24h[finalres_24h$FC > 0,])
```

```{r}
head(finalres_24h)
```

```{r}
sel_df = finalres_24h[finalres_24h$FC > 0,]

nsamples=1000
geneList = head(sel_df$FC, n=nsamples)
names(geneList) = head(sel_df$PROBEID, n=nsamples)
names(geneList) = mapIds(org.Hs.eg.db, head(sel_df$PROBEID, n=nsamples), 'ENTREZID', 'SYMBOL')
geneList = geneList[!is.na(names(geneList))]

geneList = geneList[order(geneList, decreasing = TRUE)]
head(geneList)
```

```{r}
sel_df_full = finalres_24h[finalres_24h$ADJ.PVAL < 0.05,]

geneList_full =sel_df_full$FC
names(geneList_full) = sel_df_full$PROBEID
names(geneList_full) = mapIds(org.Hs.eg.db, names(geneList_full), 'ENTREZID', 'SYMBOL')
geneList_full = geneList_full[!is.na(names(geneList_full))]

geneList_full = geneList_full[order(geneList_full, decreasing = TRUE)]
head(geneList_full)
```

```{r}
dgnGSE <- gseDGN(geneList,
              pvalueCutoff  = 0.5,
              pAdjustMethod = "BH")
dgnGSE <- setReadable(dgnGSE, 'org.Hs.eg.db')
```

```{r}
head(dgnGSE@result)
```

```{r}
dgnGSE@result[dgnGSE@result$qvalues < 0.05,]
```

```{r}
gseaplot2(dgnGSE, geneSetID = 1, title = dgnGSE$Description[1])
```

```{r}
dgnEnrich <- enrichDGN(names(geneList),
              pvalueCutoff  = 0.5,
              pAdjustMethod = "BH")
dgnEnrich <- setReadable(dgnEnrich, 'org.Hs.eg.db')
head(dgnEnrich@result, n=20)
```

```{r}
barplot(dgnEnrich, showCategory=20) 
```

```{r}
pE=mutate(dgnEnrich, qscore = -log(p.adjust, base=10)) %>%  dotplot( x="qscore", showCategory=30) + ggtitle("dotplot for ORA")
pG=mutate(dgnGSE, NESscore = NES) %>% filter(p.adjust < 0.2) %>% arrange(abs(NESscore)) %>%  dotplot( x="NESscore", showCategory=30) + ggtitle("dotplot for GSEA")

cowplot::plot_grid(pE, pG)
```

```{r}
options(repr.plot.width = 15, repr.plot.height = 10, repr.plot.res = 150)

cnetplot(dgnEnrich, categorySize="geneNum", foldChange=geneList)
```

```{r}
dgnEnrich2 <- pairwise_termsim(dgnEnrich)
treeplot(dgnEnrich2)
```

